Antigenic phenotypes and phylogenetic metrics improve forecasts of seasonal influenza A/H3N2 evolution

### 2020-06-05 manuscript/main.tex
Update bootstrap methods and results
Adds latest bootstrap p values, updates tables in main and supplement, and
updates the methods section for model comparison.

\begin{table}[htb]
  \begin{center}
    \scalebox{0.9}{
      \input{tables/simulated_model_selection.tex}
    }
    \caption{
      Model coefficients and performance on validation and test data for simulated populations ordered from best to worst by distance to the future in the validation analysis.
      Coefficients are the mean $\pm$ standard deviation for each metric in a given model across 33 training windows.
      Distance to the future (mean $\pm$ standard deviation) measures the distance in amino acids between estimated and observed future populations.
      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap hypothesis tests (see Methods).
      The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
      Test results are based on 18 timepoints not observed during model training and validation.
    }
    \label{table_simulated_model_selection}
  \end{center}
\end{table}

@@

\begin{table}[htb]
  \begin{center}
    \scalebox{0.9}{
      \input{tables/natural_model_selection.tex}
    }
    \caption{
      Model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap hypothesis tests (see Methods).
      Validation results are based on 23 timepoints.
      Test results are based on eight timepoints not observed during model training and validation.
    }
    \label{table_natural_model_selection}
  \end{center}
\end{table}

@@

\subsubsection*{Model comparison by bootstrap hypothesis testing}

We compared the performance of each biologically-informed model to the corresponding population's naive model using bootstrap hypothesis tests \cite{EfronBradley1998}.
For each model's empirical distribution of earth mover's distances to the future, $\mathbf{y}$, and the corresponding naive model's distribution, $\mathbf{x}$, we applied the following procedure to test whether their means were significantly different.
We calculated the test statistic $t$ as follows where $n$ and $m$ were the respective lengths of the naive and biologically-informed model's distributions.

$$
t = \frac{\bar{x}-\bar{y}}{\sqrt{\sigma_x^2/n + \sigma_y^2/m}}
$$

Next, we combined the two empirical distributions of distances into a single distribution, $\mathbf{z}$, and calculated the mean of this distribution.
For each original distribution, we created a new distribution that was recentered on the mean of $\mathbf{z}$.
For example, we recentered the $\mathbf{x}$ distribution as $\mathbf{x^{'}} = \mathbf{x} - \bar{x} + \bar{z}$.
We drew $n$ and $m$ samples, respectively, from the resulting distributions, $\mathbf{x^{'}}$ and $\mathbf{y^{'}}$, with replacement for 10,000 bootstrap iterations.
For each iteration, we recalculated the test statistic $t^{'}$ for the means and variances of the bootstrap samples.
We calculated the p-value as the proportion of bootstrap samples for which the $t^{'}$ statistic was greater than or equal to the observed $t$ for the empirical distributions.
To assist with the interpretation of these p-values, we further bootstrapped each model's empirical distance distribution 10,000 times, calculated the mean of each bootstrap sample, and visualized the distributions of each model annotated by its p-value from the test above (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).

>>>>

\begin{table}[htb]
  \begin{center}
    \scalebox{0.9}{
      \input{tables/simulated_model_selection.tex}
    }
    \caption{
      Model coefficients and performance on validation and test data for simulated populations ordered from best to worst by distance to the future in the validation analysis.
      Coefficients are the mean $\pm$ standard deviation for each metric in a given model across 33 training windows.
      Distance to the future (mean $\pm$ standard deviation) measures the distance in amino acids between estimated and observed future populations.
      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3}).
      The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
      Test results are based on 18 timepoints not observed during model training and validation.
    }
    \label{table_simulated_model_selection}
  \end{center}
\end{table}

@@

\begin{table}[htb]
  \begin{center}
    \scalebox{0.9}{
      \input{tables/natural_model_selection.tex}
    }
    \caption{
      Model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
      Validation results are based on 23 timepoints.
      Test results are based on eight timepoints not observed during model training and validation.
    }
    \label{table_natural_model_selection}
  \end{center}
\end{table}

@@

\subsubsection*{Model comparison by bootstrap tests}

We compared the performance of each biologically-informed model to the corresponding population's naive model using bootstrap tests.
For each model's empirical distribution of earth mover's distances to the future at a specific timepoint $t$, $y(t)$, and the corresponding naive model's distribution, $x(t)$, we calculated the difference between the two models as $y(t) - x(t)$ such that values less than zero represented timepoints when the given model outperformed the naive model.
To determine whether each model generally estimated populations that were closer to the future than the naive model, we bootstrapped the empirical difference distributions for n=10,000 samples and calculated the mean difference of each bootstrap sample (Supplemental Figs.~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3} and \ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
We calculated an empirical p value for each model as the proportion of bootstrap samples with mean values less than zero.
We identified models as significant if their empirical p value was less than 0.05.

### 2020-06-05 manuscript/supplement.tex
Update bootstrap methods and results
Adds latest bootstrap p values, updates tables in main and supplement, and
updates the methods section for model comparison.

@@

@@

\begin{table}[H]
  \begin{center}
    \scalebox{0.9}{
        \input{tables/complete_natural_model_selection.tex}
    }
    \caption{
      All model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
      Validation results are based on 23 timepoints.
      Test results are based on eight timepoints not observed during model training and validation.
      Model results for additional variants of fitness metrics including those based on epitope mutations and DMS preferences are included for reference.
    }
    \label{sup_table:complete_natural_model_selection}
  \end{center}
\end{table}

>>>>

\begin{figure}[H]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_simulated_sample_3.pdf}
  \caption{
  Bootstrap distributions of the mean difference of distances to the future between biologically-informed and naive models for simulated populations.
  Empirical differences in distances to the future were sampled with replacement and mean values for each bootstrap sample were calculated across n=10,000 bootstrap iterations.
  The horizontal gray line indicates a difference of zero between a given model and its corresponding naive model.
  Each model is annotated by the p-value representing the proportion of bootstrap samples with values less than zero (see Methods).
  }
  \label{sup_fig:bootstrap_distributions_for_simulated_sample_3}
  \end{center}
\end{figure}

@@

\begin{figure}[H]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding.pdf}
  \caption{
  Bootstrap distributions of the mean difference of distances to the future between biologically-informed and naive models for natural populations.
  Empirical differences in distances to the future were sampled with replacement and mean values for each bootstrap sample were calculated across n=10,000 bootstrap iterations.
  The horizontal gray line indicates a difference of zero between a given model and its corresponding naive model.
  Each model is annotated by the p-value representing the proportion of bootstrap samples with values less than zero (see Methods).
  }
  \label{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}
  \end{center}
\end{figure}

@@

\begin{table}[H]
  \begin{center}
    \scalebox{0.9}{
        \input{tables/complete_natural_model_selection.tex}
    }
    \caption{
      All model coefficients and performance on validation and test data for natural populations ordered from best to worst by distance to the future, as in Table~\ref{table_simulated_model_selection}.
      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Fig.~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
      Validation results are based on 23 timepoints.
      Test results are based on eight timepoints not observed during model training and validation.
      Model results for additional variants of fitness metrics including those based on epitope mutations and DMS preferences are included for reference.
    }
    \label{sup_table:complete_natural_model_selection}
  \end{center}
\end{table}
